---
title: "mappeR"
output: rmarkdown::github_document
date: "2024-09-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
devtools::load_all()
```

This is an implementation of the [mapper](https://research.math.osu.edu/tgda/mapperPBG.pdf) algorithm by Singh, Mémoli, and Carlsson, and also the [ballmapper](https://arxiv.org/pdf/1901.07410.pdf) algorithm from Dlotko.

## Setup

To install the latest version of this package from Github, run the following commands:

`install.packages("devtools")`

`library(devtools)`

`devtools::install_github("Uiowa-Applied-Topology/mappeR", upgrade=FALSE)`

`library("mappeR")`

If you're installing from Github, you might need to do some more stuff:

- **Windows:** install Rtools (http://cran.r-project.org/bin/windows/Rtools/)
- **OS X:** install Xcode (from the Mac App Store)
- **Linux:** run `apt-get install r-base-dev` (or similar).

## Examples

```{r hourglass data, echo=FALSE}
num_points = 5000

hourglass.data = data.frame( x= sapply(1:num_points, function(x) sin(x)*10) + rnorm(num_points, 0, 0.1), y = sapply(1:num_points, function(x) cos(x)*cos(x)*sin(x)*10) + rnorm(num_points, 0, 0.1), z = sapply(1:num_points, function(x) 10*sin(x)*cos(x)*sin(x))+ rnorm(num_points, 0, 0.1))

hourglass.dist = dist(hourglass.data)

eccentricity = apply(as.matrix(hourglass.dist),1,sum)/num_points

num_bins = 10
percent_overlap = 25
projx = hourglass.data$x
projy = hourglass.data$y
projz = hourglass.data$z
```

```{r plotting the curve, fig.align='center'}
library(rgl)
plot(hourglass.data, pch=20, asp=1)
plot3d(hourglass.data,pch=20)
aspect3d(1)
```

```{r mapping the mapper, fig.show="hold", out.width="50%"}

colfunc <- colorRampPalette(c("blue", "gold", "red"))
bin_colors = colfunc(num_bins)
mapperobjectx = get_1D_mapper_data(hourglass.data, projx, hourglass.dist, num_bins, percent_overlap, "single")
mapperobjecty = get_1D_mapper_data(hourglass.data, projy, hourglass.dist, num_bins, percent_overlap, "single")
mapperobjectz = get_1D_mapper_data(hourglass.data, projz, hourglass.dist, num_bins, percent_overlap, "single")

eccentricitymapperobj = get_1D_mapper_data(hourglass.data, eccentricity, hourglass.dist, num_bins, percent_overlap, "single")

coverx = create_width_balanced_cover(min(projx), max(projx), num_bins, percent_overlap)
covery = create_width_balanced_cover(min(projy), max(projy), num_bins, percent_overlap)
coverz = create_width_balanced_cover(min(projz), max(projz), num_bins, percent_overlap)
eccentricitycover = create_width_balanced_cover(min(eccentricity), max(eccentricity), num_bins, percent_overlap)

imappergraphx = imapper(mapperobjectx)
plot(imappergraphx, xlim=c(0, 2))
legend("right", title="x-coordinate", apply(coverx, 1, function(x) paste("[", round(x[1], 3), ", ", round(x[2], 3), "]")), fill=bin_colors, horiz=FALSE)

imappergraphy = imapper(mapperobjecty)
plot(imappergraphy, xlim=c(0, 2))
legend("right", title="y-coordinate", apply(covery, 1, function(x) paste("[", round(x[1], 3), ", ", round(x[2], 3), "]")), fill=bin_colors, horiz=FALSE)

imappergraphz = imapper(mapperobjectz)
plot(imappergraphz, xlim=c(0, 2))
legend("right", title="z-coordinate", apply(coverz, 1, function(x) paste("[", round(x[1], 3), ", ", round(x[2], 3), "]")), fill=bin_colors, horiz=FALSE)

iecc = imapper(eccentricitymapperobj)
plot(iecc, xlim=c(0, 2))
legend("right", title="eccentricity", apply(eccentricitycover, 1, function(x) paste("[", round(x[1], 3), ", ", round(x[2], 3), "]")), fill=bin_colors, horiz=FALSE)
```
```{r ballmapper time, fig.show='hold', out.width="50%"}
library(latex2exp)
ballmapperobj1 = get_ballmapper_data(hourglass.data, hourglass.dist, .25)
ballmapperobj2 = get_ballmapper_data(hourglass.data, hourglass.dist, .5)
ballmapperobj3 = get_ballmapper_data(hourglass.data, hourglass.dist, 1)
ballmapperobj4 = get_ballmapper_data(hourglass.data, hourglass.dist, 2)

plot(imapper(ballmapperobj1), main=TeX(r"($\epsilon=.25$)"))
plot(imapper(ballmapperobj2), main=TeX(r"($\epsilon=.5$)"))
plot(imapper(ballmapperobj3), main=TeX(r"($\epsilon=1$)"))
plot(imapper(ballmapperobj4), main=TeX(r"($\epsilon=2$)"))
```
